---
title: "Econ Thesis Data Analysis"
author: "Yeshwant Chillakuru"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# plotting
library(ggplot2)

# regressions: robust and panel
library(MASS)
library(plm)
library(foreign)
library(sandwich)
library(lmtest)
library(stargazer)

# data manipulation
library(dplyr)
library(tidyr)

# other
library(memisc)
library(pander)
```


```{r loadData, echo=FALSE}
df <- read.csv("ruralHealthAndRoadsOfIndianVillages.csv", header = TRUE, row.names = 1)
df2005 <- read.csv("ruralHealthAndRoadsOfIndianVillages2005.csv", header = TRUE, row.names = 1)
df2012 <- read.csv("ruralHealthAndRoadsOfIndianVillages2012.csv", header = TRUE, row.names = 1)
```



# I. Exploratory Data Analysis
```{r EDA1, tidy=TRUE, results='asis', include=FALSE}
dim(df)
knitr::kable(head(df))
colnames(df)
summary(df)
```

```{r EDA2, tidy=TRUE, include=TRUE, results='asis'}
colnames(df)

```

# Conduct Diff of Differences analysis
```{r EDA3, include=FALSE}

rm(dfd, diffVars, lm.diff)
diffVars <- c("id", "roadPaved", "mbTuberculosis")

dfd <- merge(df2005[, diffVars], df2012[, diffVars], by="id")

dfd[, "diffRoadPaved"] <- dfd[,"roadPaved.y"] - dfd[,"roadPaved.x"]
dfd[, "diffMbTuberculosis"] <- dfd[, "mbTuberculosis.y"] - dfd[, "mbTuberculosis.x"]

table(dfd$diffRoadPaved)
summary(dfd$diffMbTuberculosis)

lm.diff <- lm(dfd$diffMbTuberculosis ~ dfd$diffRoadPaved)
summary(lm.diff)


mydata <- df[,c("id", "year")]
mydata$time <- ifelse(mydata$year > 2005, 1, 0)



```

Between 2005 and 2012,
* `r table(dfd$diffRoadPaved)[3]` got their first paved road(s).
* `r table(dfd$diffRoadPaved)[2]` saw no change.
* `r table(dfd$diffRoadPaved)[1]` lost their paved road(s).


# Output Summary Stats for Road and Morbidity
```{r EDA4 summary states, echo=FALSE}

r2005 <- summary(df2005[, c("distanceToPavedRoad", "roadPaved")])
r2012 <- summary(df2012[, c("distanceToPavedRoad", "roadPaved")])


mbStart <- which(colnames(df) == "mbCataract")
mbEnd <- which(colnames(df) == "mbDaysIncapacitated")

mb2005 <- summary(df2005[, mbStart:mbEnd])
mb2012 <- summary(df2012[, mbStart:mbEnd])

r2005
r2012
mb2005
mb2012

rm(r2005, r2012, mb2005, mb2012)

```


------------
# II. Regressions

## 1. Univariate models
```{r univariate, echo=FALSE, tidy=TRUE}
lm.mb1 <- lm(mbDaysIncapacitated ~ roadPaved, data = df)
lm.mb2 <- lm(mbDaysIncapacitated ~ distanceToPavedRoad, data = df)

mtableUni <- mtable(
            'MB 1' = lm.mb1,
            'MB 2' = lm.mb2,
            summary.stats = c('R-squared','adj. R-squared', 'F','p','N'))

rm(lm.mb1, lm.mb2, lm.mb3)

mtableUni

rm(mtableMulti)

```


## 2. Multivariate Models
### Independent: Distance To Paved Road
```{r multivariate1, echo=FALSE}

lm.m.mb1 <- lm(mbDaysIncapacitated ~ distanceToPavedRoad + electricity + income, data=df)

lm.m.mb2 <- lm(mbDaysIncapacitated ~ distanceToPavedRoad + distanceToNearestTown + writingScore + smokeTobacco , data=df)

lm.m.mb3 <- lm(mbDaysIncapacitated ~ distanceToPavedRoad + electricity + income + distanceToNearestTown + writingScore + smokeTobacco , data=df)

mtableMulti1 <- mtable(
            'MB 1' = lm.m.mb1,
            'MB 2' = lm.m.mb2,
            'MB 3' = lm.m.mb3,
            summary.stats = c('R-squared','adj. R-squared','F','p','N'))

rm(lm.m.mb1, lm.m.mb2, lm.m.mb3)

mtableMulti1

rm(mtableMulti1)
```

### Independent: Road Paved
```{r multivariate2, echo=FALSE}

lm.m.mb1 <- lm(mbDaysIncapacitated ~ roadPaved + electricity + income, data=df)

lm.m.mb2 <- lm(mbDaysIncapacitated ~ roadPaved + distanceToNearestTown + writingScore + smokeTobacco , data=df)

lm.m.mb3 <- lm(mbDaysIncapacitated ~ roadPaved + electricity + income + distanceToNearestTown + writingScore + smokeTobacco , data=df)
lm.m.mb3.hc1 <- coeftest(lm.m.mb3, vcov = vcovHC(lm.m.mb3, "HC1"))


mtableMulti2 <- mtable(
            'MB 1' = lm.m.mb1,
            'MB 2' = lm.m.mb2,
            'MB 3' = lm.m.mb3,
            summary.stats = c('R-squared','adj. R-squared','F','p','N'))

mtableMulti2

rm(mtableMulti2)
```


```{r multivariate2OutputHTML, include=FALSE}
# comparison of different rovust vs non-robust ols
stargazer(lm.m.mb3, lm.m.mb3.hc1, type = "html", title="Results", 
          digits = 4, align = TRUE, out="robust_regression_table_comparisons.html")

openFileInOS("robust_regression_table_comparisons.html")

# clean workspace
rm( lm.m.mb1, lm.m.mb2, lm.m.mb3, lm.m.mb3.hc1)
```


## 3. Fixed Effects Regression Model
```{r fixed effects, echo=FALSE}

# distance to paved road
lm.mb1 <- lm(mbDaysIncapacitated ~ distanceToPavedRoad + electricity + income + distanceToNearestTown + writingScore + smokeTobacco, data=df)



plm.mb1 <- plm(mbDaysIncapacitated ~ distanceToPavedRoad + electricity + income + distanceToNearestTown + writingScore + smokeTobacco, data=df, index = c("id", "year"), model = "within")
plm.mb1.hc1 <- coeftest(plm.mb1, vcov = vcovHC(plm.mb1, "arellano"))
summary(plm.mb1)


# paved vs unpaved road
lm.mb2 <- lm(mbDaysIncapacitated ~ roadPaved + electricity + income + distanceToNearestTown + writingScore + smokeTobacco, data=df)


plm.mb2 <- plm(mbDaysIncapacitated ~ roadPaved + electricity + income + distanceToNearestTown + writingScore + smokeTobacco, data=df, index = c("id", "year"), model = "within")
plm.mb2.hc1 <- coeftest(plm.mb2, vcov = vcovHC(plm.mb2, "arellano"))
summary(plm.mb2)

# test if fixed effects is better. p is small, use fixed effects
pFtest(plm.mb1, lm.mb1) 
pFtest(plm.mb2, lm.mb2) 
```

```{r fixed effects OutputHTML, include=FALSE}
stargazer(plm.mb1, plm.mb1.hc1, type = "html", title="Results Fixed Effects", 
          digits = 4, align = TRUE, out="robust_regression_table_comparisons2.html")

openFileInOS("robust_regression_table_comparisons2.html")

# Clean Workspace
rm( plm.mb1, plm.mb1.hc1, plm.mb2, plm.mb2.hc1)
```


--------------------------------------------------------------------------------------------------
# Regression Sandbox

# To Do:
  * Natural Experiment of Rural Road Expansion Program --> Evan and Owens --> COPS program
      * **Can differences in differences method be used?**
      * Use roadPaved as Treatment
  * Granger Test for determining simultaneous causality?
 
 
 









