---
title: "Econ Thesis Data Analysis"
author: "Yeshwant Chillakuru"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# plotting
library(ggplot2)

# regressions: robust and panel
library(MASS)
library(plm)
library(foreign)
library(sandwich)
library(lmtest)
library(stargazer)

# data manipulation
library(dplyr)
library(tidyr)

# other
library(memisc)
library(pander)
library(knitr)
```



```{r loadData, echo=FALSE}
df <- read.csv("ruralHealthAndRoadsOfIndianVillages.csv", header = TRUE, row.names = 1)
df2005 <- read.csv("ruralHealthAndRoadsOfIndianVillages2005.csv", header = TRUE, row.names = 1)
df2012 <- read.csv("ruralHealthAndRoadsOfIndianVillages2012.csv", header = TRUE, row.names = 1)

dfNoRoad <- df[(df$roadPaved==0),]
dfNoRoad2005 <- df2005[(df2005$roadPaved==0), ]
dfNoRoad2012 <- df2012[(df2012$roadPaved==0), ]

dfNoRoad[, c('roadPaved', 'yearsWithPavedRoad')] <- NULL
dfNoRoad2005[, c('roadPaved', 'yearsWithPavedRoad')] <- NULL
dfNoRoad2012[, c('roadPaved', 'yearsWithPavedRoad')] <- NULL

dfNoRoad <- dfNoRoad[complete.cases(dfNoRoad[, c('id','year')]),]
dfNoRoad2005 <- dfNoRoad2005[complete.cases(dfNoRoad2005[, c('id','year')]),]
dfNoRoad2012 <- dfNoRoad2012[complete.cases(dfNoRoad2012[, c('id','year')]),]

dfMerge <- merge(df2005, df2012, by="id")

df$year <- as.factor(df$year)
dfNoRoad$year <- as.factor(dfNoRoad$year)

```

```{r plot, echo=FALSE, results='asis'}
ggplot(data=dfNoRoad, aes(x=distanceToPavedRoad, y=mbAnyDisease, color=year)) + geom_point(alpha=0.3, na.rm = TRUE) + labs(title = 'Disease Incidence vs Road Distance') + geom_smooth(method=lm, se=FALSE, na.rm = TRUE)
```

# Any Disease 
## Personal Controls
```{r regAnyDisease1, results='asis', echo=FALSE, message=FALSE}

plm.1 <- plm(data=df, mbAnyDisease ~ roadPaved, index = c("id", "year"), model="within")
plm.2 <- plm(data=df, mbAnyDisease ~ roadPaved + income + illiterate + smokeTobacco, index = c("id", "year"), model="within")
plm.3 <- plm(data=df, mbAnyDisease ~ roadPaved + income + illiterate + smokeTobacco + ownToilet + electricity + caste.Brahmin, index = c("id", "year"), model="within")

plm.4 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad, index = c("id", "year"), model="within")
plm.5 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + income + illiterate + smokeTobacco, index = c("id", "year"), model="within")
plm.6 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + income + illiterate + smokeTobacco + ownToilet + electricity + caste.Brahmin, index = c("id", "year"), model="within")

stargazer(plm.1, plm.2, plm.3, header=FALSE)
stargazer(plm.4, plm.5, plm.6, header=FALSE)
```

## Village Controls
```{r regAnyDisease2, results='asis', echo=FALSE, message='hide'}

plm.1 <- plm(data=df, mbAnyDisease ~ roadPaved, index = c("id", "year"), model="within")
plm.2 <- plm(data=df, mbAnyDisease ~ roadPaved + ImmunizationCampaignsNumber + drinkingWaterSource.Piped, index = c("id", "year"), model="within")
plm.3 <- plm(data=df, mbAnyDisease ~ roadPaved + healthSubCenter + primaryHealthCenter + communityHealthCenter, index = c("id", "year"), model="within")
plm.4 <- plm(data=df, mbAnyDisease ~ roadPaved + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + healthSubCenter + primaryHealthCenter + communityHealthCenter, index = c("id", "year"), model="within")

plm.5 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad, index = c("id", "year"), model="within")
plm.6 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + ImmunizationCampaignsNumber + drinkingWaterSource.Piped, index = c("id", "year"), model="within")
plm.7 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + healthSubCenter + primaryHealthCenter + communityHealthCenter, index = c("id", "year"), model="within")
plm.8 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + healthSubCenter + primaryHealthCenter + communityHealthCenter, index = c("id", "year"), model="within")

stargazer(plm.2, plm.3, plm.4, header=FALSE)
stargazer(plm.6, plm.7, plm.8, header=FALSE)

```

## Medical TreatmentWhere and TreatmentWho Controls
```{r regAnyDisease3, results='asis', echo=FALSE, message=FALSE}
plm.1 <- plm(data=df, mbAnyDisease ~ roadPaved, index = c("id", "year"), model="within")
plm.2 <- plm(data=df, mbAnyDisease ~ roadPaved + mbTreatmentWho1.PublicDoc + mbTreatmentWho1.PublicDocInPvt + mbTreatmentWho1.PvtDoc + mbTreatmentWho1.Pharm, index = c("id", "year"), model="within")
plm.3 <- plm(data=df, mbAnyDisease ~ roadPaved + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.4 <- plm(data=df, mbAnyDisease ~ roadPaved + mbTreatmentWho1.PublicDoc + mbTreatmentWho1.PublicDocInPvt + mbTreatmentWho1.PvtDoc + mbTreatmentWho1.Pharm + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")

plm.5 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad, index = c("id", "year"), model="within")
plm.6 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + mbTreatmentWho1.PublicDoc + mbTreatmentWho1.PublicDocInPvt + mbTreatmentWho1.PvtDoc + mbTreatmentWho1.Pharm, index = c("id", "year"), model="within")
plm.7 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.8 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + mbTreatmentWho1.PublicDoc + mbTreatmentWho1.PublicDocInPvt + mbTreatmentWho1.PvtDoc + mbTreatmentWho1.Pharm + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")

stargazer(plm.2, plm.3, plm.4, header=FALSE)
stargazer(plm.6, plm.7, plm.8, header=FALSE)
```

## Combined Controls
```{r regAnyDisease4, results='asis', echo=FALSE, message=FALSE}
plm.1 <- plm(data=df, mbAnyDisease ~ roadPaved, index = c("id", "year"), model="within")
plm.2 <- plm(data=df, mbAnyDisease ~ roadPaved + income + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.3 <- plm(data=df, mbAnyDisease ~ roadPaved + illiterate + smokeTobacco + ownToilet + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.4 <- plm(data=df, mbAnyDisease ~ roadPaved + income + illiterate + smokeTobacco + ownToilet + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")

plm.5 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad, index = c("id", "year"), model="within")
plm.6 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + income + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.7 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + illiterate + smokeTobacco + ownToilet + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.8 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + income + illiterate + smokeTobacco + ownToilet + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")


stargazer(plm.2, plm.3, plm.4, header=FALSE)
stargazer(plm.6, plm.7, plm.8, header=FALSE)
```

# Communicable Disease

```{r regComDisease1, results='asis', echo=FALSE, message=FALSE}
plm.1 <- plm(data=df, mbComDisease ~ roadPaved, index = c("id", "year"), model="within")
plm.2 <- plm(data=df, mbComDisease ~ roadPaved + income + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.3 <- plm(data=df, mbComDisease ~ roadPaved + illiterate + smokeTobacco + ownToilet + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.4 <- plm(data=df, mbComDisease ~ roadPaved + income + illiterate + smokeTobacco + ownToilet + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")

plm.5 <- plm(data=dfNoRoad, mbComDisease ~ lnDistanceToPavedRoad, index = c("id", "year"), model="within")
plm.6 <- plm(data=dfNoRoad, mbComDisease ~ lnDistanceToPavedRoad + income + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.7 <- plm(data=dfNoRoad, mbComDisease ~ lnDistanceToPavedRoad + illiterate + smokeTobacco + ownToilet + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.8 <- plm(data=dfNoRoad, mbComDisease ~ lnDistanceToPavedRoad + income + illiterate + smokeTobacco + ownToilet + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")


stargazer(plm.2, plm.3, plm.4, header=FALSE)
stargazer(plm.6, plm.7, plm.8, header=FALSE)
```

# Non Communicable Disease
```{r regNonComDisease1, results='asis', echo=FALSE, message=FALSE}
plm.1 <- plm(data=df, mbNonComDisease ~ roadPaved, index = c("id", "year"), model="within")
plm.2 <- plm(data=df, mbNonComDisease ~ roadPaved + income + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.3 <- plm(data=df, mbNonComDisease ~ roadPaved + illiterate + smokeTobacco + ownToilet + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.4 <- plm(data=df, mbNonComDisease ~ roadPaved + income + illiterate + smokeTobacco + ownToilet + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")

plm.5 <- plm(data=dfNoRoad, mbNonComDisease ~ lnDistanceToPavedRoad, index = c("id", "year"), model="within")
plm.6 <- plm(data=dfNoRoad, mbNonComDisease ~ lnDistanceToPavedRoad + income + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.7 <- plm(data=dfNoRoad, mbNonComDisease ~ lnDistanceToPavedRoad + illiterate + smokeTobacco + ownToilet + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.8 <- plm(data=dfNoRoad, mbNonComDisease ~ lnDistanceToPavedRoad + income + illiterate + smokeTobacco + ownToilet + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")


stargazer(plm.2, plm.3, plm.4, header=FALSE)
stargazer(plm.6, plm.7, plm.8, header=FALSE)
```

# STD or AIDS
```{r regSTDorAIDS1, results='asis', echo=FALSE, message=FALSE}
plm.1 <- plm(data=df, mbSTDorAIDS ~ roadPaved, index = c("id", "year"), model="within")
plm.2 <- plm(data=df, mbSTDorAIDS ~ roadPaved + income + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.3 <- plm(data=df, mbSTDorAIDS ~ roadPaved + illiterate + smokeTobacco + ownToilet + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.4 <- plm(data=df, mbSTDorAIDS ~ roadPaved + income + illiterate + smokeTobacco + ownToilet + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")


plm.5 <- plm(data=dfNoRoad, mbSTDorAIDS ~ lnDistanceToPavedRoad, index = c("id", "year"), model="within")
plm.6 <- plm(data=dfNoRoad, mbSTDorAIDS ~ lnDistanceToPavedRoad + income + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.7 <- plm(data=dfNoRoad, mbSTDorAIDS ~ lnDistanceToPavedRoad + illiterate + smokeTobacco + ownToilet + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")
plm.8 <- plm(data=dfNoRoad, mbSTDorAIDS ~ lnDistanceToPavedRoad + income + illiterate + smokeTobacco + ownToilet + ImmunizationCampaignsNumber + drinkingWaterSource.Piped + primaryHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model="within")


stargazer(plm.2, plm.3, plm.4, header=FALSE)
stargazer(plm.6, plm.7, plm.8, header=FALSE)
```



--------------------------------------------------------------------------------------------------
# Notes
  * Regression
    * Controls:
        * Need to control for how often individuals go to hospital (more people sick in 2005 then in 2012 --> increased diagnosis?)
  

# To Do:
  * 
  
  
# Questions
  * If using state-fixed and time-fixed effects for each village, do I really need all these controls?
  * Can I do a differences-in-differences? <- potentially do for paved vs unpaved
  * What controls should I include and what shouldn't? Should I be worried about "controlling away" the actual effect?
    
 
 
 
 
```{r everythingRegression, include=FALSE}
# EVERYTHING Regression
lm <- lm(data=df, mbAnyDisease ~ roadPaved + ImmunizationCampaignsNumber + mbTreatmentRecieved + mbCostOfTreatment + mbTravelExpenses + smokeTobacco + chewTobacco + drinkAlcohol + illiterate + attendedSchool + income + ownToilet + electricity + healthSubCenter + primaryHealthCenter + communityHealthCenter + govtMaternityCenter + pvtClinicTrainedDoc + pvtClinicUntrainedDoc + pvtHospital + pvtPharm + otherGovtMedFacility + distHealthSubCenter + distPrimaryHealthCenter + distCommunityHealthCenter + distDistrictHospital + distGovtMaternityCenter + distPvtClinicTrainedDoc + distPvtClinicUntrainedDoc + distPvtHospital + distPvtPharm + distOtherGovtMedFacility + caste.Brahmin + caste.OBC + caste.SC + caste.ST + drinkingWaterSource.Piped + drinkingWaterSource.TubeWell + drinkingWaterSource.HandPump + drinkingWaterSource.OpenWell + drinkingWaterSource.CoveredWell + drinkingWaterSource.River + drinkingWaterSource.Pond + drinkingWaterSource.Truck + drinkingWaterSource.RainWater + drinkingWaterSource.Bottled + drinkingWaterSource.Other + mbTreatmentWho1.NA + mbTreatmentWho1.PublicDoc + mbTreatmentWho1.PublicDocInPvt + mbTreatmentWho1.PvtDoc + mbTreatmentWho1.Pharm + mbTreatmentWhere1.NA + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown + mbTreatmentWho2.NA + mbTreatmentWho2.PublicDoc + mbTreatmentWho2.PublicDocInPvt + mbTreatmentWho2.PvtDoc + mbTreatmentWho2.Pharm + mbTreatmentWhere2.NA + mbTreatmentWhere2.SameVillage + mbTreatmentWhere2.AnotherVillage + mbTreatmentWhere2.OtherTown + mbTreatmentWhere2.DistrictTown)
lm.hc1 <- coeftest(lm, vcov=vcovHC(lm, "HC1"))
```









