---
title: "Econ Thesis Data Analysis"
author: "Yeshwant Chillakuru"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# plotting
library(ggplot2)

# regressions: robust and panel
library(MASS)
library(plm)
library(foreign)
library(sandwich)
library(lmtest)
library(stargazer)

# data manipulation
library(dplyr)
library(tidyr)

# other
library(memisc)
library(pander)
library(knitr)
```

# Load Data
```{r loadData, echo=FALSE}
df <- read.csv("ruralHealthAndRoadsOfIndianVillages.csv", header = TRUE, row.names = 1)
df2005 <- read.csv("ruralHealthAndRoadsOfIndianVillages2005.csv", header = TRUE, row.names = 1)
df2012 <- read.csv("ruralHealthAndRoadsOfIndianVillages2012.csv", header = TRUE, row.names = 1)

dfNoRoad <- df[(df$roadPaved==0),]
dfNoRoad2005 <- df2005[(df2005$roadPaved==0), ]
dfNoRoad2012 <- df2012[(df2012$roadPaved==0), ]

dfMerge <- merge(df2005, df2012, by="id")

df$year <- as.factor(df$year)

```

# 3. Data
## Dimensions of dataset
  * 2005 -> `r dim(df2005)`
  * 2012 -> `r dim(df2012)`
  * Number of Villages Same in 2005 and 2012 -> `r dim(dfMerge)[1]` villages

## Road Summary Statistics
```{r dataRoad, include=FALSE}

pctPvdRd05 <- mean(df2005$roadPaved, na.rm = TRUE)*100
pctPvdRd12 <- mean(df2012$roadPaved, na.rm = TRUE)*100

avgDistRd2005 <- mean(dfNoRoad2005$distanceToPavedRoad, na.rm = TRUE)
avgDistRd2012 <- mean(dfNoRoad2012$distanceToPavedRoad, na.rm = TRUE)


rm(temp_dfRoad2005, temp_dfRoad2012)
```

Percent of villages with a paved road:

  * 2005 -> `r pctPvdRd05`%
  * 2012 -> `r pctPvdRd12`%
  
Average distance to road for villages without roads:

  * 2005 -> `r avgDistRd2005` km
  * 2012 -> `r avgDistRd2012` km
  
## Health Summary Statistics
```{r dataHealth, include=FALSE, message=FALSE}

avg_mbAnyDisease05 <- mean(df2005$mbAnyDisease, na.rm=TRUE)
avg_mbAnyDisease12 <- mean(df2012$mbAnyDisease, na.rm=TRUE)

avg_mbComDisease05 <- mean(df2005$mbComDisease, na.rm=TRUE)
avg_mbComDisease12 <- mean(df2012$mbComDisease, na.rm=TRUE)

avg_mbNonComDisease05 <- mean(df2005$mbNonComDisease, na.rm=TRUE)
avg_mbNonComDisease12 <- mean(df2012$mbNonComDisease, na.rm=TRUE)

diseaseMeansDF <- data.frame(Year=c('2005', '2012'))
startDis <- which(colnames(df) == 'mbCataract')
endDis <- which(colnames(df) == 'mbOtherLongTerm')
diseaseNames <- colnames(df[,startDis:endDis])
for(disease in diseaseNames){
  diseaseMeansDF[,disease] <- c(mean(df2005[,disease]), mean(df2012[,disease]))
}


rm(startDis, endDis, disease)
```

Percent Sick with Any Disease:

  * 2005 -> `r avg_mbAnyDisease05`%
  * 2012 -> `r avg_mbAnyDisease12`%
  
Percent Sick with Communicable Disease:

  * 2005 -> `r avg_mbComDisease05`%
  * 2012 -> `r avg_mbComDisease12`%
  
Percent Sick with Non-communicable Disease:

  * 2005 -> `r avg_mbNonComDisease05`%
  * 2012 -> `r avg_mbNonComDisease12`%
  
Percent Sick with Non-communicable Disease:

  * 2005 -> `r avg_mbNonComDisease05`%
  * 2012 -> `r avg_mbNonComDisease12`%
  
  
```{r dataTables, echo=FALSE, results='asis'}
kable(diseaseMeansDF[,1:7])
kable(diseaseMeansDF[,c(1,8:14)])
```


```{r dataPlot, echo=FALSE, message=FALSE}
ggplot(data=df, aes(x=distanceToPavedRoad, y=mbAnyDisease, color=year)) + geom_point(alpha=0.3, na.rm = TRUE) + labs(title = 'Disease Incidence vs Road Distance') + geom_smooth(method=lm, se=FALSE, na.rm = TRUE)
```


# Regressions
$$diseaseIncidence_{it} = \beta RoadPaved_{it} + \delta_{1} Ind_{it} + \delta_{2} Household_{it} + \delta_{3} Village_{it} + \varepsilon_{it}$$

$$diseaseIncidence_{it} = \beta DistanceToPavedRoad_{it} + \delta_{1} Ind_{it} + \delta_{2} Household_{it} + \delta_{3} Village_{it} + \varepsilon_{it}$$

## Any Disease

### Paved vs Unpaved
```{r regAnyDisease1, echo=FALSE, results='asis', message=FALSE}

#univariate
lm.1 <- lm(data=df, mbAnyDisease ~ roadPaved)
lm.1.hc1 <- coeftest(lm.1, vcov=vcovHC(lm.1, "HC1"))

plm.1 <- plm(data=df, mbAnyDisease ~ roadPaved, index = c("id", "year"), model = "within")
plm.2 <- plm(data=df, mbAnyDisease ~ roadPaved + ImmunizationCampaignsNumber + healthSubCenter + primaryHealthCenter + communityHealthCenter, index = c("id", "year"), model = "within")
plm.3 <- plm(data=df, mbAnyDisease ~ roadPaved + smokeTobacco + illiterate + ownToilet + caste.Brahmin + caste.OBC + caste.SC + caste.ST, index = c("id", "year"), model = "within")
plm.4 <- plm(data=df, mbAnyDisease ~ roadPaved + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")
plm.5 <- plm(data=df, mbAnyDisease ~ roadPaved + ImmunizationCampaignsNumber + smokeTobacco + illiterate + ownToilet + healthSubCenter + primaryHealthCenter + communityHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")

stargazer(plm.1, plm.2, plm.3, header=FALSE)
stargazer(plm.4, plm.5, header=FALSE)
# stargazer(plm.1, plm.2, plm.3, plm.4, plm.5)
```

### Distance to Paved for villages without Road
```{r regAnyDisease2, echo=FALSE, results='asis', message=FALSE}

#univariate
lm.1 <- lm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad)
lm.1.hc1 <- coeftest(lm.1, vcov=vcovHC(lm.1, "HC1"))

plm.1 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad, index = c("id", "year"), model = "within")
plm.2 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + ImmunizationCampaignsNumber + healthSubCenter + primaryHealthCenter + communityHealthCenter, index = c("id", "year"), model = "within")
plm.3 <- plm(data=df, mbAnyDisease ~ lnDistanceToPavedRoad + smokeTobacco + illiterate + ownToilet + caste.Brahmin + caste.OBC + caste.SC + caste.ST, index = c("id", "year"), model = "within")
plm.4 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")
plm.5 <- plm(data=dfNoRoad, mbAnyDisease ~ lnDistanceToPavedRoad + ImmunizationCampaignsNumber + smokeTobacco + illiterate + ownToilet + healthSubCenter + primaryHealthCenter + communityHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")

stargazer(plm.1, plm.2, plm.3, header=FALSE)
stargazer(plm.4, plm.5, header=FALSE)
# stargazer(plm.1, plm.2, plm.3, plm.4, plm.5, type="text")
```

## Communicable Disease
### Paved vs Unpaved
```{r regComDisease1, echo=FALSE, results='asis', message=FALSE}

#univariate
lm.1 <- lm(data=df, mbComDisease ~ roadPaved)
lm.1.hc1 <- coeftest(lm.1, vcov=vcovHC(lm.1, "HC1"))

plm.1 <- plm(data=df, mbComDisease ~ roadPaved, index = c("id", "year"), model = "within")
plm.2 <- plm(data=df, mbComDisease ~ roadPaved + ImmunizationCampaignsNumber + healthSubCenter + primaryHealthCenter + communityHealthCenter, index = c("id", "year"), model = "within")
plm.3 <- plm(data=df, mbComDisease ~ roadPaved + smokeTobacco + illiterate + ownToilet + caste.Brahmin + caste.OBC + caste.SC + caste.ST, index = c("id", "year"), model = "within")
plm.4 <- plm(data=df, mbComDisease ~ roadPaved + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")
plm.5 <- plm(data=df, mbComDisease ~ roadPaved + ImmunizationCampaignsNumber + smokeTobacco + illiterate + ownToilet + healthSubCenter + primaryHealthCenter + communityHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")

stargazer(plm.1, plm.2, plm.3, header=FALSE)
stargazer(plm.4, plm.5, header=FALSE)
# stargazer(plm.1, plm.2, plm.3, plm.4, plm.5)
```

### Distance to Paved for villages without Road
```{r regComDisease2, echo=FALSE, results='asis', message=FALSE}

#univariate
lm.1 <- lm(data=dfNoRoad, mbComDisease ~ lnDistanceToPavedRoad)
lm.1.hc1 <- coeftest(lm.1, vcov=vcovHC(lm.1, "HC1"))

plm.1 <- plm(data=dfNoRoad, mbComDisease ~ lnDistanceToPavedRoad, index = c("id", "year"), model = "within")
plm.2 <- plm(data=dfNoRoad, mbComDisease ~ lnDistanceToPavedRoad + ImmunizationCampaignsNumber + healthSubCenter + primaryHealthCenter + communityHealthCenter, index = c("id", "year"), model = "within")
plm.3 <- plm(data=df, mbComDisease ~ lnDistanceToPavedRoad + smokeTobacco + illiterate + ownToilet + caste.Brahmin + caste.OBC + caste.SC + caste.ST, index = c("id", "year"), model = "within")
plm.4 <- plm(data=dfNoRoad, mbComDisease ~ lnDistanceToPavedRoad + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")
plm.5 <- plm(data=dfNoRoad, mbComDisease ~ lnDistanceToPavedRoad + ImmunizationCampaignsNumber + smokeTobacco + illiterate + ownToilet + healthSubCenter + primaryHealthCenter + communityHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")

stargazer(plm.1, plm.2, plm.3, header=FALSE)
stargazer(plm.4, plm.5, header=FALSE)
# stargazer(plm.1, plm.2, plm.3, plm.4, plm.5, type="text")
```


## Non-Communicable Disease
### Paved vs Unpaved
```{r regNonComDisease1, echo=FALSE, results='asis', message=FALSE}

#univariate
lm.1 <- lm(data=df, mbNonComDisease ~ roadPaved)
lm.1.hc1 <- coeftest(lm.1, vcov=vcovHC(lm.1, "HC1"))

plm.1 <- plm(data=df, mbNonComDisease ~ roadPaved, index = c("id", "year"), model = "within")
plm.2 <- plm(data=df, mbNonComDisease ~ roadPaved + ImmunizationCampaignsNumber + healthSubCenter + primaryHealthCenter + communityHealthCenter, index = c("id", "year"), model = "within")
plm.3 <- plm(data=df, mbNonComDisease ~ roadPaved + smokeTobacco + illiterate + ownToilet + caste.Brahmin + caste.OBC + caste.SC + caste.ST, index = c("id", "year"), model = "within")
plm.4 <- plm(data=df, mbNonComDisease ~ roadPaved + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")
plm.5 <- plm(data=df, mbNonComDisease ~ roadPaved + ImmunizationCampaignsNumber + smokeTobacco + illiterate + ownToilet + healthSubCenter + primaryHealthCenter + communityHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")

plm.6 <- plm(data=df, mbNonComDisease ~ roadPaved + mbTreatmentWho1.PublicDoc + mbTreatmentWho1.PublicDocInPvt + mbTreatmentWho1.PvtDoc + mbTreatmentWho1.Pharm, index = c("id", "year"), model = "within")
summary(plm.6)


stargazer(plm.1, plm.2, plm.3, header=FALSE)
stargazer(plm.4, plm.5, header=FALSE)
# stargazer(plm.1, plm.2, plm.3, plm.4, plm.5)
```

### Distance to Paved for villages without Road
```{r regNonComDisease2, echo=FALSE, results='asis', message=FALSE}

#univariate
lm.1 <- lm(data=dfNoRoad, mbNonComDisease ~ lnDistanceToPavedRoad)
lm.1.hc1 <- coeftest(lm.1, vcov=vcovHC(lm.1, "HC1"))

plm.1 <- plm(data=dfNoRoad, mbNonComDisease ~ lnDistanceToPavedRoad, index = c("id", "year"), model = "within")
plm.2 <- plm(data=dfNoRoad, mbNonComDisease ~ lnDistanceToPavedRoad + ImmunizationCampaignsNumber + healthSubCenter + primaryHealthCenter + communityHealthCenter, index = c("id", "year"), model = "within")
plm.3 <- plm(data=df, mbNonComDisease ~ lnDistanceToPavedRoad + smokeTobacco + illiterate + ownToilet + caste.Brahmin + caste.OBC + caste.SC + caste.ST, index = c("id", "year"), model = "within")
plm.4 <- plm(data=dfNoRoad, mbNonComDisease ~ lnDistanceToPavedRoad + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")
plm.5 <- plm(data=dfNoRoad, mbNonComDisease ~ lnDistanceToPavedRoad + ImmunizationCampaignsNumber + smokeTobacco + illiterate + ownToilet + healthSubCenter + primaryHealthCenter + communityHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")

stargazer(plm.1, plm.2, plm.3, header=FALSE)
stargazer(plm.4, plm.5, header=FALSE)
# stargazer(plm.1, plm.2, plm.3, plm.4, plm.5, type="text")
```

## STDs or AIDS
### Paved vs Unpaved
```{r regSTDDisease1, echo=FALSE, results='asis', message=FALSE}

#univariate
lm.1 <- lm(data=df, mbSTDorAIDS ~ roadPaved)
lm.1.hc1 <- coeftest(lm.1, vcov=vcovHC(lm.1, "HC1"))

plm.1 <- plm(data=df, mbSTDorAIDS ~ roadPaved, index = c("id", "year"), model = "within")
plm.2 <- plm(data=df, mbSTDorAIDS ~ roadPaved + ImmunizationCampaignsNumber + healthSubCenter + primaryHealthCenter + communityHealthCenter, index = c("id", "year"), model = "within")
plm.3 <- plm(data=df, mbSTDorAIDS ~ roadPaved + smokeTobacco + illiterate + ownToilet + caste.Brahmin + caste.OBC + caste.SC + caste.ST, index = c("id", "year"), model = "within")
plm.4 <- plm(data=df, mbSTDorAIDS ~ roadPaved + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")
plm.5 <- plm(data=df, mbSTDorAIDS ~ roadPaved + ImmunizationCampaignsNumber + smokeTobacco + illiterate + ownToilet + healthSubCenter + primaryHealthCenter + communityHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")

stargazer(plm.1, plm.2, plm.3, header=FALSE)
stargazer(plm.4, plm.5, header=FALSE)
# stargazer(plm.1, plm.2, plm.3, plm.4, plm.5)
```

### Distance to Paved for villages without Road
```{r regSTDDisease2, echo=FALSE, results='asis', message=FALSE}

#univariate
lm.1 <- lm(data=dfNoRoad, mbSTDorAIDS ~ lnDistanceToPavedRoad)
lm.1.hc1 <- coeftest(lm.1, vcov=vcovHC(lm.1, "HC1"))

plm.1 <- plm(data=dfNoRoad, mbSTDorAIDS ~ lnDistanceToPavedRoad, index = c("id", "year"), model = "within")
plm.2 <- plm(data=dfNoRoad, mbSTDorAIDS ~ lnDistanceToPavedRoad + ImmunizationCampaignsNumber + healthSubCenter + primaryHealthCenter + communityHealthCenter, index = c("id", "year"), model = "within")
plm.3 <- plm(data=df, mbSTDorAIDS ~ lnDistanceToPavedRoad + smokeTobacco + illiterate + ownToilet + caste.Brahmin + caste.OBC + caste.SC + caste.ST, index = c("id", "year"), model = "within")
plm.4 <- plm(data=dfNoRoad, mbSTDorAIDS ~ lnDistanceToPavedRoad + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")
plm.5 <- plm(data=dfNoRoad, mbSTDorAIDS ~ lnDistanceToPavedRoad + ImmunizationCampaignsNumber + smokeTobacco + illiterate + ownToilet + healthSubCenter + primaryHealthCenter + communityHealthCenter + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown, index = c("id", "year"), model = "within")

stargazer(plm.1, plm.2, plm.3, header=FALSE)
stargazer(plm.4, plm.5, header=FALSE)
# stargazer(plm.1, plm.2, plm.3, plm.4, plm.5, type="text")
```





--------------------------------------------------------------------------------------------------
# Notes
  * Regression
    * Controls:
        * Need to control for how often individuals go to hospital (more people sick in 2005 then in 2012 --> increased diagnosis?)
      
# Mention in paper
   * Income had no impact with village and time fixed effects

# To Do:
  * Natural Experiment of Rural Road Expansion Program --> Evan and Owens --> COPS program
      * **Can differences in differences method be used?** 
      * Use roadPaved as Treatment
  * Granger Test for determining simultaneous causality?
  
  
# Questions
  * If using state-fixed and time-fixed effects for each village, do I really need all these controls?
  * Can I do a differences-in-differences? <- potentially do for paved vs unpaved
  * What controls should I include and what shouldn't? Should I be worried about "controlling away" the actual effect?
    
 
 
 
 
```{r everythingRegression, include=FALSE}
# EVERYTHING Regression
lm <- lm(data=df, mbAnyDisease ~ roadPaved + ImmunizationCampaignsNumber + mbTreatmentRecieved + mbCostOfTreatment + mbTravelExpenses + smokeTobacco + chewTobacco + drinkAlcohol + illiterate + attendedSchool + income + ownToilet + electricity + healthSubCenter + primaryHealthCenter + communityHealthCenter + govtMaternityCenter + pvtClinicTrainedDoc + pvtClinicUntrainedDoc + pvtHospital + pvtPharm + otherGovtMedFacility + distHealthSubCenter + distPrimaryHealthCenter + distCommunityHealthCenter + distDistrictHospital + distGovtMaternityCenter + distPvtClinicTrainedDoc + distPvtClinicUntrainedDoc + distPvtHospital + distPvtPharm + distOtherGovtMedFacility + caste.Brahmin + caste.OBC + caste.SC + caste.ST + drinkingWaterSource.Piped + drinkingWaterSource.TubeWell + drinkingWaterSource.HandPump + drinkingWaterSource.OpenWell + drinkingWaterSource.CoveredWell + drinkingWaterSource.River + drinkingWaterSource.Pond + drinkingWaterSource.Truck + drinkingWaterSource.RainWater + drinkingWaterSource.Bottled + drinkingWaterSource.Other + mbTreatmentWho1.NA + mbTreatmentWho1.PublicDoc + mbTreatmentWho1.PublicDocInPvt + mbTreatmentWho1.PvtDoc + mbTreatmentWho1.Pharm + mbTreatmentWhere1.NA + mbTreatmentWhere1.SameVillage + mbTreatmentWhere1.AnotherVillage + mbTreatmentWhere1.OtherTown + mbTreatmentWhere1.DistrictTown + mbTreatmentWho2.NA + mbTreatmentWho2.PublicDoc + mbTreatmentWho2.PublicDocInPvt + mbTreatmentWho2.PvtDoc + mbTreatmentWho2.Pharm + mbTreatmentWhere2.NA + mbTreatmentWhere2.SameVillage + mbTreatmentWhere2.AnotherVillage + mbTreatmentWhere2.OtherTown + mbTreatmentWhere2.DistrictTown)
lm.hc1 <- coeftest(lm, vcov=vcovHC(lm, "HC1"))
```









